---
title: "Bias and Power of Slope Versus Final-Visit Estimands
  Under Nonlinear Treatment Effects in Alzheimer's Disease
  Trials"
author: "Ronald G. Thomas, Ph.D.\\thanks{Wertheim School of Public Health, UCSD; ORCID: 0000-0003-1686-4965}"
date: '`r format(Sys.Date(), "%B %d, %Y")`'
output:
  bookdown::pdf_document2:
    latex_engine: xelatex
    toc: true
    toc_depth: 2
    number_sections: true
    keep_tex: true
    highlight: tango
fontsize: 11pt
linestretch: 1.1
geometry: 'left=3cm,right=3cm,top=2.5cm,bottom=2.5cm'
bibliography: references.bib
csl: statistics-in-medicine.csl
link-citations: true
header-includes:
  - \usepackage{amsmath}
  - \usepackage{amssymb}
  - \usepackage{booktabs}
  - \usepackage{longtable}
  - \usepackage{float}
  - \usepackage[mathlines]{lineno}
  - \linenumbers
  - \usepackage{placeins}
knit: (function(input, ...) { d <- dirname(input); while (!file.exists(file.path(d, 'tools', 'stamp-render.R')) && d != dirname(d)) d <- dirname(d); source(file.path(d, 'tools', 'stamp-render.R'))$value(input) })
---

```{r setup, include=FALSE}
options(
  knitr.kable.NA = "",
  dplyr.summarise.inform = FALSE,
  knitr.table.format = "latex"
)
library(tidyverse)
library(knitr)
library(kableExtra)
library(MASS)
library(nlme)
library(mmrm)

opts_chunk$set(
  warning = FALSE,
  message = FALSE,
  echo = FALSE,
  fig.width = 5,
  fig.height = 3.5,
  results = "asis",
  dev = "pdf",
  cache = TRUE,
  cache.path = "cache/"
)
options(scipen = 1, digits = 3)
```

```{r load-sim-08, include=FALSE}
# Loaded here (before the Abstract) so its dynamic inline values are
# available throughout the document, including the Abstract. See
# Methods, "Simulation procedure", and the Reproducibility statement
# for the exact commands used to generate these cached files.
source("../scripts/sim_mmrm_linearity.R")

sim_08_path <- "../data/derived_data/sim_08.rds"
if (file.exists(sim_08_path)) {
  sim_08 <- readRDS(sim_08_path)
  sim_raw <- sim_08$res
  sim_summary <- sim_08$summary
} else {
  warning(
    "analysis/data/derived_data/sim_08.rds not found; falling back ",
    "to a live, single-core rerun. Run analysis/scripts/run_sim_08.R ",
    "to regenerate the cached file used for the reported results."
  )
  RNGkind("L'Ecuyer-CMRG")
  set.seed(20260310)
  sim_raw <- run_simulation(
    n_rep     = 1000,
    kappa_vec = c(0.3, 0.5, 0.7, 1.0, 1.5, 2.0, 3.0),
    n_per_arm = 200,
    delta     = -0.45
  )
  sim_summary <- summarize_results(sim_raw, true_delta = -0.45)
}

sim_08_null_path <- "../data/derived_data/sim_08_null.rds"
sim_null_available <- file.exists(sim_08_null_path)
if (sim_null_available) {
  sim_08_null <- readRDS(sim_08_null_path)
  sim_null_summary <- sim_08_null$summary
}

abstract_linear_row <- sim_summary |> dplyr::filter(kappa == 1.0)
abstract_worst_slope_bias <- sim_summary |>
  dplyr::slice_max(abs(slope_bias), n = 1, with_ties = FALSE)
abstract_cat_cov_range <- range(sim_summary$cat_coverage, na.rm = TRUE)
abstract_slope_cov_range <- range(sim_summary$slope_coverage, na.rm = TRUE)
abstract_gap_closed <- abstract_linear_row$slope_c_power >=
  abstract_linear_row$cat_power
```

# Abstract {-}

We use Monte Carlo simulation to quantify how nonlinearity in
the treatment effect trajectory affects the bias, coverage,
and power of a random-slopes mixed model relative to a
categorical-time MMRM, in the setting of an 18-month
Alzheimer's disease trial with CDR-SB as the endpoint and a
fixed total treatment effect of $-0.45$ points at month 18
(matching CLARITY-AD). Across seven values of a shape
parameter $\kappa$ spanning early-onset to delayed-onset
treatment effect trajectories, the random-slopes model is
approximately unbiased only near $\kappa = 1$ (exact
linearity) and its bias grows to
`r sprintf('%.3f', abs(abstract_worst_slope_bias$slope_bias))`
points (versus a true effect of $-0.45$) at the most convex
trajectory examined ($\kappa = `r abstract_worst_slope_bias$kappa`$);
its 95\% CI coverage falls as low as
`r sprintf('%.3f', abstract_slope_cov_range[1])` in the same
condition. The categorical-time MMRM remains approximately
unbiased and maintains coverage at or above
`r sprintf('%.3f', abstract_cat_cov_range[1])` across all
conditions examined. Contrary to the conventional expectation
that a correctly specified parsimonious mean structure is more
efficient, the unconstrained random-slopes model does not
achieve higher power than the categorical-time MMRM even under
exact linearity in this design; we show this is attributable in
part to an estimator choice (an unidentified treatment main
effect retained in a model fit to post-baseline data only)
rather than to the linearity assumption per se. A full-scale
($n_{\mathrm{rep}} = 1000$ per $\kappa$) evaluation of a
constrained alternative that removes this main effect
`r if (abstract_gap_closed) "closes the power gap with the categorical-time MMRM under exact linearity" else "narrows, but does not fully close, the power gap with the categorical-time MMRM under exact linearity"`
(constrained power
`r sprintf('%.3f', abstract_linear_row$slope_c_power)` versus
`r sprintf('%.3f', abstract_linear_row$cat_power)` at
$\kappa = 1$), while a separate null-scenario simulation
confirms all three estimators control type I error near the
nominal 0.05 level.

**Keywords:** mixed model for repeated measures; random
slopes; Alzheimer's disease; CDR-SB; estimand; simulation
study; nonlinearity.

# Data and code availability {-}

The simulation driver (`analysis/scripts/sim_mmrm_linearity.R`),
the report source (`analysis/report/report.Rmd`), and this
compendium's `renv.lock` are version-controlled in the
`mmrmrobust` repository. `analysis/scripts/run_sim_08.R`
reproduces the cached simulation output loaded by this report
(`analysis/data/derived_data/sim_08.rds`); see the
Reproducibility statement at the end of this document for the
exact command and RNG settings used.

# Introduction

## The CDR-SB as a primary endpoint in AD trials

The Clinical Dementia Rating Sum of Boxes (CDR-SB) has
become the primary efficacy endpoint for disease-modifying
trials in early Alzheimer's disease (AD). The instrument
integrates cognition and function into a single composite
score ranging from 0 to 18, and it demonstrates superior
responsiveness to change in early-stage populations compared
with traditional cognitive scales such as the ADAS-Cog
[@cedarbaum2013rationale; @coley2011suitability]. A parallel
line of work has instead built composite cognitive endpoints
for the preclinical stage, such as the Preclinical Alzheimer
Cognitive Composite [@donohue2014pacc]; CDR-SB remains the
dominant choice once clinical impairment is present, which
is the setting considered here. Annual
CDR-SB progression rates in placebo arms of recent trials
are approximately 1.3 to 1.7 points per year for
participants with baseline CDR global scores of 0.5 to 1.0
[@cedarbaum2013rationale; @hartz2025clinical].

## Treatment effects from recent anti-amyloid trials

The approval of lecanemab and the positive results for
donanemab have provided the first evidence that
amyloid-targeting therapies can slow clinical decline on
CDR-SB. In the CLARITY-AD trial, lecanemab reduced 18-month
CDR-SB worsening by 0.45 points (27% slowing) relative to
placebo, with the treatment arm declining 1.21 points
versus 1.66 in placebo [@vandyck2023lecanemab]. The
TRAILBLAZER-ALZ 2 trial of donanemab demonstrated a similar
magnitude of slowing on CDR-SB in participants with
low/medium tau burden [@sims2023donanemab]. By contrast, the
earlier Phase 3 studies of aducanumab did not show a
consistent CDR-SB benefit across trials
[@budd2022aducanumab], underscoring that both the magnitude
and the shape of the treatment-effect trajectory are
empirical questions rather than assured features of
amyloid-targeting therapy. In the two positive trials,
inspection of the CDR-SB trajectories over time reveals a
non-constant treatment difference: the separation between
arms widens gradually, with the rate of separation
appearing to change across the 18-month follow-up period.

## The linearity question

Both regulatory agencies and the statistical community have
debated the choice between two estimands for these trials:

- **Final-visit estimand:** The between-group difference in
  mean change from baseline at the last scheduled visit
  (e.g., month 18). This corresponds to the categorical-time
  MMRM, which places no parametric constraint on how the mean
  response evolves over time
  [@mallinckrodt2008recommendations].

- **Slope estimand:** The between-group difference in the
  rate of change per unit time. This corresponds to a
  random-slopes mixed model that assumes the treatment
  effect accumulates linearly over the follow-up period.

The U.S. Food and Drug Administration's 2018 draft guidance
on early AD drug development discusses the use of slope
analyses as a potentially more sensitive approach for
detecting disease-modifying effects, particularly when the
mechanism is expected to produce a constant rate of slowing
[@fda2018guidance]. The European Medicines Agency's guideline
similarly acknowledges slope-based endpoints for
disease-modification claims, while also noting that the
assumption of constant treatment effect rates must be
verified [@ema2018guideline]. The ICH E9(R1) addendum on
estimands provides the formal framework for distinguishing
these two targets of estimation, emphasizing that the choice
of estimand should precede the choice of statistical model
[@ich2019e9r1].

## Why linearity may not hold

Several biological and design factors suggest that the
treatment effect in AD trials is unlikely to be perfectly
linear in time:

1. **Delayed pharmacodynamic onset.** Anti-amyloid
   antibodies require weeks to months to achieve meaningful
   amyloid clearance. The clinical benefit may lag behind
   the biological effect, producing a delayed-onset or
   concave treatment effect trajectory (the effect
   accelerates over time).

2. **Plateau or diminishing returns.** Conversely, if the
   treatment primarily benefits patients at earlier disease
   stages, the marginal benefit per additional month of
   treatment could decrease as disease progresses,
   producing a convex (early-onset) trajectory.

3. **Nonlinear placebo progression.** Even if the treatment
   modifies the underlying rate by a constant fraction,
   nonlinearity in the natural disease trajectory
   (documented by @jamalian2023modeling using nonlinear
   mixed-effect models) would produce a nonlinear absolute
   treatment difference over time.

Despite the clinical relevance of this question, limited
simulation evidence exists quantifying how much nonlinearity
the random-slopes model can tolerate. @chen2018simulation
compared the slope model with the categorical-time MMRM for
ADAS-Cog data, finding that the slope model is more powerful
when linearity approximately holds, but biased otherwise.
That study, however, did not systematically vary the degree
of nonlinearity, nor did it focus on CDR-SB.

## Present study

We use Monte Carlo simulation to evaluate the
robustness of a random-slopes mixed model (a linear
mixed-effects model, not itself an MMRM in the clinical-trials
sense; see Methods) to departures from linearity in the
treatment effect trajectory. We simulate 18-month AD trials
with CDR-SB as the primary endpoint, visits at 3-month
intervals, and a total treatment effect of $-0.45$ points at
month 18 (matching CLARITY-AD). We parameterize the shape of
the treatment effect trajectory using a power function
$g(t) = (t/t_{\max})^\kappa$, where $\kappa = 1$ yields a
linear effect, $\kappa < 1$ yields an early-onset (convex)
pattern, and $\kappa > 1$ yields a delayed-onset (concave)
pattern. We compare three analysis models:

1. An **unconstrained random-slopes model** that fits
   treatment-by-time as a linear interaction (retaining a
   treatment main effect) and extracts the treatment contrast
   at 18 months from the interaction term alone.
2. A **constrained random-slopes model** that omits the
   treatment main effect, imposing equality of the two arms at
   baseline.
3. A **categorical-time MMRM** that fits treatment-by-visit
   as an unstructured interaction and extracts the treatment
   contrast at the final visit.

We evaluate bias, empirical standard error, model-based
standard error, power (proportion of $p < 0.05$), and 95\%
confidence interval coverage across a range of $\kappa$
values, and additionally report empirical type I error for all
three models in a separate null-scenario ($\delta = 0$)
simulation.

# Methods

## ADEMP structure

Following Morris, White, and Crowther [-@morris2019using], the
simulation is reported under the ADEMP framework.

- **Aims.** Quantify the robustness of a random-slopes model fit
  (which imposes linear-in-time treatment effects) to nonlinear
  treatment trajectories governed by the shape parameter $\kappa$.
- **Data-generating mechanism.** A longitudinal model where the
  treatment effect accrues as $g(t, \kappa) = (t / t_{\max})^\kappa$,
  with $\kappa$ varied across $\{0.3, 0.5, 0.7, 1.0, 1.5, 2.0, 3.0\}$.
- **Estimand.** The total treatment effect at month 18,
  $\delta = -0.45$, which is invariant to $\kappa$ by construction.
- **Methods.** (i) Unconstrained random-slopes model via
  \texttt{nlme::lme}, which assumes a linear-in-time treatment
  effect and retains a treatment main effect; (ii) constrained
  random-slopes model via \texttt{nlme::lme}, omitting the
  treatment main effect; (iii) categorical-time MMRM via the
  \texttt{mmrm} package with unstructured covariance.
- **Performance measures.** Convergence rate, bias, empirical SE,
  mean model-based SE, power at $\alpha = 0.05$, and coverage of 95\%
  CIs, each reported with its Monte Carlo SE per Morris Table 6; a
  separate null-scenario ($\delta = 0$) simulation additionally
  reports empirical type I error for all three models.

## Data-generating model

We generate data for $N = 200$ participants per arm (total
$N = 400$) observed at months $t \in \{0, 3, 6, 9, 12,
15, 18\}$. The outcome for participant $i$ at visit $j$ is:

$$
Y_{ij} = \beta_0 + b_{0i} + (\beta_1 + b_{1i}) t_j^* +
\delta \cdot g(t_j, \kappa) \cdot \text{trt}_i +
\varepsilon_{ij}
$$

where $t_j^* = t_j / 12$ is time in years, $\text{trt}_i$
is a binary treatment indicator, and:

$$
g(t_j, \kappa) = \left(\frac{t_j}{t_{\max}}\right)^\kappa
$$

The random effects $(b_{0i}, b_{1i})$ follow a bivariate
normal distribution with standard deviations $\sigma_{b_0}
= 1.0$ and $\sigma_{b_1} = 0.6$ (in CDR-SB points per
year) and correlation $\rho = 0.3$. The within-subject
errors $\varepsilon_{ij}$ have an AR(1) structure with
marginal variance $\sigma^2 = 1.44$ and autocorrelation
$\phi = 0.4$.

## Parameter values

The fixed-effect parameters are calibrated to approximate
the placebo arm of CLARITY-AD:

| Parameter | Value | Interpretation |
|-----------|-------|----------------|
| $\beta_0$ | 3.2 | Mean baseline CDR-SB |
| $\beta_1$ | 1.5 points/year | Placebo annual decline |
| $\delta$ | $-0.45$ | Treatment effect at $t_{\max}$ |
| $\kappa$ | Varies | Shape of treatment effect |

The $\kappa$ parameter takes values in $\{0.3, 0.5, 0.7,
1.0, 1.5, 2.0, 3.0\}$, spanning strongly convex
(early-onset) to strongly concave (delayed-onset) treatment
effect trajectories.

## Analysis models

Both models are fit to post-baseline observations only
(months 3 through 18); neither model adjusts for baseline
CDR-SB, which departs from the near-universal practice of
baseline-adjusted change-from-baseline MMRM
[@mallinckrodt2008recommendations]. We disclose this as a
design choice made to keep the two models on comparable
footing (a categorical model with visit 3 as the reference
level already absorbs baseline-like information from the
first post-randomization visit); it is not evaluated as a
sensitivity condition here and is noted as a limitation
below.

**Random-slopes model (unconstrained).** We fit a linear
mixed-effects model (via `nlme::lme`) with fixed effects for
treatment, time (continuous, in years), and their
interaction, plus random intercepts and random slopes for
time. Because the fixed-effect specification includes a
treatment main effect while the model is fit to post-baseline
data only (excluding $t = 0$, where the true treatment effect
is 0 by randomization), the main effect is not anchored to an
observed baseline difference. The treatment effect at 18
months is extracted as $\hat{\beta}_{\text{trt} \times
\text{time}} \times 1.5$ only, discarding the fitted main
effect; this is an incomplete model-implied contrast rather
than the full linear combination
$\hat{\beta}_{\text{trt}} + \hat{\beta}_{\text{trt} \times
\text{time}} \times 1.5$. As shown in the Results and
discussed further below, this estimator choice, not
nonlinearity per se, is a plausible driver of the model's
power deficit at $\kappa = 1$.

**Random-slopes model (constrained).** As a remediation, we
also implement a constrained variant that omits the treatment
main effect entirely (`y ~ time + trt:time`, random intercept
and slope by subject), imposing equality of the two arms at
baseline as in constrained longitudinal data analysis (cLDA)
[@liang2000sankhya; @liu2015slopes]. This is standard practice
for slope estimands in the trials literature and removes the
unidentified main effect implicated above. We report the
full-scale evaluation of this estimator ($n_{\mathrm{rep}} =
1000$ per $\kappa$, all seven $\kappa$ values) in Results,
"Estimator definition check", and Table 1.

**Categorical-time MMRM.** We fit the standard marginal MMRM
used in clinical trials (via `mmrm::mmrm`) with fixed effects
for treatment, visit (categorical), and their interaction, and
a fully unstructured covariance matrix (heterogeneous
variances and unstructured correlations across visits). This
is the model recommended by
@mallinckrodt2008recommendations, in the general treatment of
longitudinal data analysis by @liang1986longitudinal and
@fitzmaurice2011applied, and implemented in the `mmrm` R
package [@sabanesbove2024mmrm]. The treatment effect
at month 18 is estimated as the linear combination of the
main treatment effect and the treatment-by-month-18
interaction; its standard error and $p$-value are obtained
from a Satterthwaite-adjusted one-degree-of-freedom contrast
(`mmrm::df_1d`), not the delta method.

## Simulation procedure

For each value of $\kappa$, we generate 1000 replicate
trials, fit all three models (unconstrained random-slopes,
constrained random-slopes, categorical-time MMRM) to each, and
record the point estimate, standard error, and $p$-value for the
18-month treatment contrast. Summary metrics are computed across
replicates. For power MCSE $\leq 1.1$ pp at $p = 0.5$,
$n_{\mathrm{rep}} \geq 2{,}066$; $n_{\mathrm{rep}} = 1000$ gives
power MCSE $\approx 1.58$ pp, adequate for initial iteration but
should be raised for publication. $N = 200$ per arm is a
working design point rather than a value calibrated to a
target power for this specific comparison; power calculations
for AD trials more generally are discussed by @ard2011.
CLARITY-AD itself randomized approximately 900 participants
per arm [@vandyck2023lecanemab], so the power estimates below
should not be read as representative of that trial's operating
characteristics (see Limitations).

`sim_raw`, `sim_summary`, and (if available) `sim_null_summary`
were loaded from the cached simulation output at the top of this
document (chunk `load-sim-08`), before the Abstract, so that
dynamic figures are available throughout the manuscript; see the
Reproducibility statement for the exact regeneration commands and
RNG settings.

# Results

## Headline findings

```{r headline-08, include=FALSE}
# Uses `sim_summary` produced by the `run-sim` chunk.
linear_row <- sim_summary |> dplyr::filter(kappa == 1.0)
worst_slope_bias <- sim_summary |>
  dplyr::slice_max(abs(slope_bias), n = 1, with_ties = FALSE)
worst_cat_bias <- sim_summary |>
  dplyr::slice_max(abs(cat_bias), n = 1, with_ties = FALSE)
slope_cov_range <- range(sim_summary$slope_coverage, na.rm = TRUE)
cat_cov_range <- range(sim_summary$cat_coverage, na.rm = TRUE)
slope_pow_range <- range(sim_summary$slope_power, na.rm = TRUE)
cat_pow_range <- range(sim_summary$cat_power, na.rm = TRUE)
```

We observe that under linear treatment accrual ($\kappa = 1$),
both the random-slopes model and the categorical-time MMRM
recover the true effect with negligible bias
(`r sprintf('%.3f', linear_row$slope_bias)` and
`r sprintf('%.3f', linear_row$cat_bias)` respectively,
MCSE `r sprintf('%.3f', linear_row$mcse_slope_bias)` and
`r sprintf('%.3f', linear_row$mcse_cat_bias)`). Away from
linearity, however, the random-slopes model's bias grows
sharply and its power and coverage are worse than the
categorical-time MMRM's at every $\kappa$ examined, including
$\kappa = 1$: at $\kappa = 1$ the random-slopes model's power
is `r sprintf('%.3f', linear_row$slope_power)` versus
`r sprintf('%.3f', linear_row$cat_power)` for the
categorical-time MMRM, and its empirical SE
(`r sprintf('%.3f', linear_row$slope_emp_se)`) exceeds the
categorical-time MMRM's
(`r sprintf('%.3f', linear_row$cat_emp_se)`). This contradicts
the usual expectation that a correctly specified, more
parsimonious mean structure is more efficient, and is
addressed directly in the "Estimator definition check"
subsection below and in the Discussion. Across all
`r nrow(sim_summary)` values of $\kappa \in
\{0.3, 0.5, 0.7, 1, 1.5, 2, 3\}$, the random-slopes
estimator's largest absolute bias was
`r sprintf('%.3f', abs(worst_slope_bias$slope_bias))` at
$\kappa = `r worst_slope_bias$kappa`$, while the
categorical-time MMRM never exceeded
`r sprintf('%.3f', abs(worst_cat_bias$cat_bias))`.
Coverage of the random-slopes confidence interval ranged
from `r sprintf('%.3f', slope_cov_range[1])` to
`r sprintf('%.3f', slope_cov_range[2])`; the
categorical-time MMRM maintained coverage in
[`r sprintf('%.3f', cat_cov_range[1])`,
`r sprintf('%.3f', cat_cov_range[2])`] across the same
range. Power for the random-slopes model ranged from
`r sprintf('%.3f', slope_pow_range[1])` to
`r sprintf('%.3f', slope_pow_range[2])`, and for the
categorical-time MMRM from
`r sprintf('%.3f', cat_pow_range[1])` to
`r sprintf('%.3f', cat_pow_range[2])`; the categorical-time
model's power is higher, or effectively tied, at every
$\kappa$. Full Morris Table 6 MCSEs and convergence rates are
reported in Table 1.

## Treatment effect shape

Figure 1 illustrates the treatment effect trajectories
$g(t, \kappa)$ for the seven values of $\kappa$ used in the
simulation. When $\kappa = 1$, the treatment effect accrues
linearly. When $\kappa < 1$ (e.g., $\kappa = 0.3$), most
of the effect is realized early, with diminishing marginal
benefit at later visits. When $\kappa > 1$ (e.g., $\kappa
= 3.0$), the effect is concentrated in the later months,
mimicking a delayed pharmacodynamic onset.

```{r fig-shapes, fig.cap="Treatment effect trajectories under different shape parameters. All curves reach the same total effect at month 18.", fig.width=5, fig.height=3.5}
t_months <- seq(0, 18, by = 0.5)
kappa_vals <- c(0.3, 0.5, 0.7, 1.0, 1.5, 2.0, 3.0)
shape_dat <- expand.grid(
  t = t_months,
  kappa = kappa_vals
) |>
  mutate(
    g = (t / 18)^kappa,
    effect = -0.45 * g,
    kappa_label = factor(
      paste0("kappa = ", kappa),
      levels = paste0("kappa = ", kappa_vals)
    )
  )

ltype_map <- c(
  "kappa = 0.3" = "dotted",
  "kappa = 0.5" = "dashed",
  "kappa = 0.7" = "dotdash",
  "kappa = 1"   = "solid",
  "kappa = 1.5" = "longdash",
  "kappa = 2"   = "twodash",
  "kappa = 3"   = "4212"
)

ggplot(shape_dat, aes(
  x = t, y = effect,
  linetype = kappa_label
)) +
  geom_line(linewidth = 0.6) +
  scale_linetype_manual(values = ltype_map) +
  labs(
    x = "Month",
    y = "Treatment effect (CDR-SB points)",
    linetype = "Shape"
  ) +
  theme_minimal(base_size = 10) +
  theme(legend.position = "right")
```

## Summary statistics

Table 1 presents the full simulation results across all
seven values of $\kappa$.

\FloatBarrier

```{r tab-summary}
tab_display <- sim_summary |>
  dplyr::select(
    kappa,
    slope_bias, slope_emp_se, slope_mean_se,
    slope_power, slope_coverage,
    slope_c_bias, slope_c_emp_se, slope_c_mean_se,
    slope_c_power, slope_c_coverage,
    cat_bias, cat_emp_se, cat_mean_se,
    cat_power, cat_coverage
  ) |>
  mutate(across(
    c(slope_bias, slope_c_bias, cat_bias),
    ~ round(.x, 3)
  )) |>
  mutate(across(
    c(slope_emp_se, slope_mean_se,
      slope_c_emp_se, slope_c_mean_se,
      cat_emp_se, cat_mean_se),
    ~ round(.x, 3)
  )) |>
  mutate(across(
    c(slope_power, slope_c_power, cat_power,
      slope_coverage, slope_c_coverage, cat_coverage),
    ~ round(.x, 3)
  ))

kbl(
  tab_display,
  format = "latex",
  booktabs = TRUE,
  escape = FALSE,
  caption = paste(
    "Simulation results: bias, empirical SE, mean",
    "model-based SE, power, and 95\\% CI coverage",
    "for the unconstrained random-slopes model, the constrained",
    "random-slopes model, and the categorical-time MMRM",
    "across shape parameter values ($n_{\\mathrm{rep}} = 1000$",
    "per $\\kappa$). True treatment effect = $-0.45$."
  ),
  col.names = c(
    "$\\kappa$",
    "Bias", "Emp SE", "Mod SE", "Power", "Cov",
    "Bias", "Emp SE", "Mod SE", "Power", "Cov",
    "Bias", "Emp SE", "Mod SE", "Power", "Cov"
  )
) |>
  add_header_above(c(
    " " = 1,
    "Random-Slopes (Unconstrained)" = 5,
    "Random-Slopes (Constrained)" = 5,
    "Categorical-Time MMRM" = 5
  )) |>
  kable_styling(
    latex_options = c("striped", "scale_down"),
    font_size = 7
  )
```

Table 2 reports the convergence rate for each model and the
Monte Carlo standard errors (MCSEs) for bias and power, per
Morris, White, and Crowther [-@morris2019using] Table 6; MCSEs
for the model-based SE and coverage are reported in
`analysis/data/derived_data/sim_08.rds` but omitted from the
printed table for space.

\FloatBarrier

```{r tab-mcse}
tab_mcse <- sim_summary |>
  dplyr::select(
    kappa,
    slope_conv, mcse_slope_bias, mcse_slope_power,
    slope_c_conv, mcse_slope_c_bias, mcse_slope_c_power,
    cat_conv, mcse_cat_bias, mcse_cat_power
  ) |>
  mutate(across(
    c(slope_conv, slope_c_conv, cat_conv),
    ~ round(.x, 3)
  )) |>
  mutate(across(
    c(mcse_slope_bias, mcse_slope_power,
      mcse_slope_c_bias, mcse_slope_c_power,
      mcse_cat_bias, mcse_cat_power),
    ~ round(.x, 4)
  ))

kbl(
  tab_mcse,
  format = "latex",
  booktabs = TRUE,
  escape = FALSE,
  caption = paste(
    "Convergence rates and Monte Carlo standard errors (MCSE)",
    "for bias and power, per Morris, White, and Crowther",
    "(2019) Table 6."
  ),
  col.names = c(
    "$\\kappa$",
    "Conv.", "MCSE(bias)", "MCSE(power)",
    "Conv.", "MCSE(bias)", "MCSE(power)",
    "Conv.", "MCSE(bias)", "MCSE(power)"
  )
) |>
  add_header_above(c(
    " " = 1,
    "Random-Slopes (Unconstrained)" = 3,
    "Random-Slopes (Constrained)" = 3,
    "Categorical-Time MMRM" = 3
  )) |>
  kable_styling(
    latex_options = c("striped", "scale_down"),
    font_size = 8
  )
```

## Bias

Figure 2 displays the bias of each model as a function of
$\kappa$. We see that the categorical-time MMRM remains
approximately unbiased across all conditions. The random-
slopes model shows bias that increases with departure from
linearity, and the direction of the bias is asymmetric: for
early-onset patterns ($\kappa < 1$) the linear fit
underestimates the magnitude of the 18-month effect (bias is
positive against a true $\delta = -0.45$, i.e. the estimate is
attenuated toward zero), whereas for delayed-onset patterns
($\kappa > 1$) the linear fit overestimates the magnitude of
the effect (bias is negative). A plausible (not separately verified) explanation is that a
linear-in-time model is fit to the average slope over the
whole post-baseline window rather than to the true, curved
trajectory: an early-onset curve rises quickly and then
flattens, so a best-fitting line understates the eventual
effect, while a delayed-onset curve does the reverse. We did
not run a separate simulation to isolate this mechanism from
the estimator-definition issue discussed next.

\FloatBarrier

```{r fig-bias, fig.cap="Bias in the estimated 18-month treatment effect as a function of the shape parameter. Both random-slopes variants show increasing bias as the trajectory departs from linearity, while the categorical-time model remains approximately unbiased.", fig.width=5, fig.height=3}
bias_dat <- sim_summary |>
  dplyr::select(kappa, slope_bias, slope_c_bias, cat_bias) |>
  pivot_longer(
    cols = c(slope_bias, slope_c_bias, cat_bias),
    names_to = "model",
    values_to = "bias"
  ) |>
  mutate(model = dplyr::case_match(
    model,
    "slope_bias" ~ "Random slopes (unconstrained)",
    "slope_c_bias" ~ "Random slopes (constrained)",
    "cat_bias" ~ "Categorical time"
  ))

ggplot(bias_dat, aes(
  x = kappa, y = bias,
  linetype = model, shape = model
)) +
  geom_hline(yintercept = 0, linewidth = 0.3,
             color = "grey50") +
  geom_line(linewidth = 0.6) +
  geom_point(size = 2) +
  scale_linetype_manual(values = c(
    "Random slopes (unconstrained)" = "solid",
    "Random slopes (constrained)" = "dotted",
    "Categorical time" = "dashed"
  )) +
  scale_shape_manual(values = c(
    "Random slopes (unconstrained)" = 16,
    "Random slopes (constrained)" = 15,
    "Categorical time" = 17
  )) +
  labs(
    x = expression(kappa),
    y = "Bias (CDR-SB points)",
    linetype = "Model",
    shape = "Model"
  ) +
  theme_minimal(base_size = 10) +
  theme(legend.position = "bottom", legend.direction = "vertical")
```

## Power comparison

Figure 3 compares the power of all three models. The
categorical-time MMRM has higher power than the unconstrained
random-slopes model at every $\kappa$ examined, including
$\kappa = 1$ (exact linearity), where a correctly specified
parametric model would conventionally be expected to be the
more efficient choice. This is the opposite of the qualitative
pattern the unconstrained model's relative simplicity would
suggest, and the "Estimator definition check" subsection below
shows that the way the 18-month contrast is extracted from the
random-slopes model, not linearity itself, is a plausible
driver of this deficit at $\kappa = 1$; the constrained
random-slopes model's power, also shown in Figure 3, is
reported and discussed there. Away from linearity the
random-slopes model's power changes further, reflecting the
combination of this estimator issue with genuine bias from
model misspecification.

## Estimator definition check

```{r estimator-check-08, include=FALSE}
slope_c_gap_closed <- linear_row$slope_c_power >= linear_row$cat_power
slope_c_power_gain <- linear_row$slope_c_power - linear_row$slope_power
slope_c_bias_dir_txt <- if (worst_slope_bias$slope_c_bias * worst_slope_bias$slope_bias < 0) {
  "reversed the sign of"
} else if (abs(worst_slope_bias$slope_c_bias) < abs(worst_slope_bias$slope_bias)) {
  "reduced the magnitude of, without reversing the sign of,"
} else {
  "did not reduce"
}
```

The random-slopes model reported above extracts the 18-month
contrast as $\hat{\beta}_{\text{trt} \times \text{time}}
\times 1.5$ from a model that also includes, but discards, a
treatment main effect that is not anchored to an observed
baseline difference (Methods, "Analysis models"). To test
whether this estimator choice, rather than linearity, explains
the power deficit at $\kappa = 1$, we ran the full-scale design
($n_{\mathrm{rep}} = 1000$ per $\kappa$, all seven $\kappa$
values) comparing the unconstrained estimator above against the
constrained estimator described in Methods (no treatment main
effect, `y ~ time + trt:time`); results appear as the middle
block of Table 1 and Table 2. At $\kappa = 1$ (exact linearity),
the constrained model's power was
`r sprintf('%.3f', linear_row$slope_c_power)`, compared with
`r sprintf('%.3f', linear_row$slope_power)` for the unconstrained
model and `r sprintf('%.3f', linear_row$cat_power)` for the
categorical-time MMRM
(`r if (slope_c_gap_closed) "matching or exceeding" else "still below"`
the categorical-time model's power, a change of
`r sprintf('%+.3f', slope_c_power_gain)` power points relative to
the unconstrained estimator). Its empirical SE at $\kappa = 1$ was
`r sprintf('%.3f', linear_row$slope_c_emp_se)`, versus
`r sprintf('%.3f', linear_row$slope_emp_se)` for the unconstrained
model and `r sprintf('%.3f', linear_row$cat_emp_se)` for the
categorical-time MMRM. Away from linearity, at the most convex
condition examined
($\kappa = `r worst_slope_bias$kappa`$), the constrained
estimator's bias was `r sprintf('%.3f', worst_slope_bias$slope_c_bias)`
against the unconstrained model's
`r sprintf('%.3f', worst_slope_bias$slope_bias)` (true
$\delta = -0.45$); the constrained estimator
`r slope_c_bias_dir_txt` the unconstrained model's bias in this
condition. Both random-slopes variants converged at rates
reported in Table 2. These full-scale results supersede the
small-scale, $n_{\mathrm{rep}} = 20$ preliminary check reported in
an earlier revision of this manuscript.

## Type I error under the null

```{r null-08, include=FALSE}
if (!sim_null_available) {
  stop(
    "analysis/data/derived_data/sim_08_null.rds not found; run ",
    "analysis/scripts/run_sim_08_null.R before rendering this report."
  )
}
null_row <- sim_null_summary |> dplyr::filter(kappa == 1.0)
```

Power comparisons between models are difficult to interpret without
also confirming that each model controls type I error at the nominal
level (pub_review whitepaper 2026-08-16, major issue 3). We
simulated a null scenario ($\delta = 0$; the shape parameter $\kappa$
has no effect when $\delta = 0$, so a single, nominal $\kappa = 1$
condition suffices) with $n_{\mathrm{rep}} = 1000$ replicates and
otherwise identical design parameters to the main simulation, and
recorded the empirical rejection rate at $\alpha = 0.05$ for all
three models -- the empirical type I error rate under this design.
The unconstrained random-slopes model's empirical type I error was
`r sprintf('%.4f', null_row$slope_power)` (MCSE
`r sprintf('%.4f', null_row$mcse_slope_power)`), the constrained
random-slopes model's was `r sprintf('%.4f', null_row$slope_c_power)`
(MCSE `r sprintf('%.4f', null_row$mcse_slope_c_power)`), and the
categorical-time MMRM's was `r sprintf('%.4f', null_row$cat_power)`
(MCSE `r sprintf('%.4f', null_row$mcse_cat_power)`), against a
nominal level of 0.05.

```{r tab-null}
tab_null <- null_row |>
  dplyr::select(
    slope_conv, slope_power, mcse_slope_power,
    slope_c_conv, slope_c_power, mcse_slope_c_power,
    cat_conv, cat_power, mcse_cat_power
  ) |>
  mutate(across(everything(), ~ round(.x, 4)))

kbl(
  tab_null,
  format = "latex",
  booktabs = TRUE,
  escape = FALSE,
  caption = paste(
    "Empirical type I error rate at $\\alpha = 0.05$",
    "($\\delta = 0$, $n_{\\mathrm{rep}} = 1000$) for all three",
    "models, with Monte Carlo standard errors."
  ),
  col.names = c(
    "Conv.", "Type I error", "MCSE",
    "Conv.", "Type I error", "MCSE",
    "Conv.", "Type I error", "MCSE"
  )
) |>
  add_header_above(c(
    "Random-Slopes (Unconstrained)" = 3,
    "Random-Slopes (Constrained)" = 3,
    "Categorical-Time MMRM" = 3
  )) |>
  kable_styling(
    latex_options = c("striped", "scale_down"),
    font_size = 8
  )
```

\FloatBarrier

```{r fig-power, fig.cap="Power to detect the 18-month treatment effect ($\\alpha = 0.05$, two-sided) as a function of the shape parameter.", fig.width=5, fig.height=3}
power_dat <- sim_summary |>
  dplyr::select(kappa, slope_power, slope_c_power, cat_power) |>
  pivot_longer(
    cols = c(slope_power, slope_c_power, cat_power),
    names_to = "model",
    values_to = "power"
  ) |>
  mutate(model = dplyr::case_match(
    model,
    "slope_power" ~ "Random slopes (unconstrained)",
    "slope_c_power" ~ "Random slopes (constrained)",
    "cat_power" ~ "Categorical time"
  ))

ggplot(power_dat, aes(
  x = kappa, y = power,
  linetype = model, shape = model
)) +
  geom_line(linewidth = 0.6) +
  geom_point(size = 2) +
  scale_linetype_manual(values = c(
    "Random slopes (unconstrained)" = "solid",
    "Random slopes (constrained)" = "dotted",
    "Categorical time" = "dashed"
  )) +
  scale_shape_manual(values = c(
    "Random slopes (unconstrained)" = 16,
    "Random slopes (constrained)" = 15,
    "Categorical time" = 17
  )) +
  labs(
    x = expression(kappa),
    y = "Power",
    linetype = "Model",
    shape = "Model"
  ) +
  theme_minimal(base_size = 10) +
  theme(legend.position = "bottom", legend.direction = "vertical")
```

\FloatBarrier

# Discussion

In this study, we quantify the bias, coverage, and power of
three analysis models (unconstrained random-slopes, constrained
random-slopes, and categorical-time MMRM) when the treatment
effect in an AD trial departs from linearity, and find that the
unconstrained random-slopes model performs worse than the
categorical-time MMRM even under exact linearity ($\kappa = 1$):
it has lower power
(`r sprintf('%.3f', linear_row$slope_power)` vs.
`r sprintf('%.3f', linear_row$cat_power)`), a larger empirical SE
(`r sprintf('%.3f', linear_row$slope_emp_se)` vs.
`r sprintf('%.3f', linear_row$cat_emp_se)`), and, away from
linearity, materially larger bias and lower coverage than the
categorical-time model, which remains approximately unbiased
with coverage at or above
`r sprintf('%.3f', cat_cov_range[1])` across every condition
examined. This contradicts the general
principle that a correctly specified, more parsimonious mean
structure should be more efficient than a saturated one
[@mallinckrodt2008recommendations], and it also contradicts a
qualitative expectation from @chen2018simulation's comparison
of the two model classes for ADAS-Cog.

As $\kappa$ departs from 1, the random-slopes model incurs
additional, asymmetric bias in estimating the 18-month
treatment contrast (Results, "Bias"), consistent with a
linear-in-time model averaging over a curved trajectory. The
categorical-time model, by contrast, places no constraint on
the shape of the mean trajectory and remains approximately
unbiased regardless of $\kappa$.

The power deficit at $\kappa = 1$ cannot be explained by
nonlinearity, since there is none in that condition, and we
attribute it instead to how the random-slopes model's 18-month
contrast is constructed (Methods, "Analysis models"; Results,
"Estimator definition check"): the model retains a treatment
main effect that is not anchored to an observed baseline
difference (post-baseline-only data, randomized baseline), and
the reported contrast discards that main effect rather than
using it or omitting it from the model. Our full-scale check
($n_{\mathrm{rep}} = 1000$ per $\kappa$, all seven $\kappa$
values) of a constrained estimator that omits the main effect
[@liang2000sankhya; @liu2015slopes]
`r if (slope_c_gap_closed) "closed the power gap with the categorical-time MMRM at $\\kappa = 1$" else "narrowed, but did not close, the power gap with the categorical-time MMRM at $\\kappa = 1$"`
(constrained-model power
`r sprintf('%.3f', linear_row$slope_c_power)` versus
`r sprintf('%.3f', linear_row$cat_power)` for the categorical-time
MMRM and `r sprintf('%.3f', linear_row$slope_power)` for the
unconstrained model), and at the most convex condition examined
($\kappa = `r worst_slope_bias$kappa`$) the constrained
estimator `r slope_c_bias_dir_txt` the unconstrained model's
bias (`r sprintf('%.3f', worst_slope_bias$slope_c_bias)` versus
`r sprintf('%.3f', worst_slope_bias$slope_bias)`, true
$\delta = -0.45$). We also confirmed, in a separate null-scenario
simulation ($\delta = 0$, $n_{\mathrm{rep}} = 1000$; Results,
"Type I error under the null"), that all three estimators
control type I error near the nominal 0.05 level, so the power
differences reported here are not an artifact of differential
false-positive rates.
`r if (slope_c_gap_closed) "This is consistent with the hypothesis that the unidentified, unanchored treatment main effect -- not linearity per se -- drove the unconstrained model's power deficit under exact linearity." else "The gap did not fully close, so the unidentified main effect appears to be only a partial explanation of the unconstrained model's power deficit under exact linearity; some residual disadvantage for the random-slopes specification (constrained or not) relative to the categorical-time MMRM remains unexplained by this study."`

The practical implication we draw for AD trial design is a
conditional one: the choice between slope and final-visit
estimands should be informed by the expected treatment-effect
trajectory, but only after confirming that the slope estimand is
implemented with an estimator that does not itself introduce a
power penalty under linearity, such as the constrained variant
evaluated here.
`r if (slope_c_gap_closed) "When there is strong biological rationale for a constant rate of slowing (e.g., a treatment that directly modulates the underlying disease rate) and the constrained estimator is used, the slope estimand and random-slopes model may offer a power advantage on par with the categorical-time MMRM." else "Even with the constrained estimator, this simulation did not find the random-slopes model's power to match the categorical-time MMRM's under exact linearity, so we do not recommend the slope estimand as a power-superior alternative to the final-visit estimand in this design, regardless of estimator, absent further investigation of the residual gap."`
When the mechanism suggests delayed onset (as with
amyloid-clearing antibodies that require months to achieve
plaque reduction) or early plateau, the final-visit estimand
with categorical-time MMRM is the safer choice regardless of
estimator, because it remains approximately unbiased with
coverage at or above `r sprintf('%.3f', min(sim_summary$cat_coverage, na.rm = TRUE))`
across every $\kappa$ examined, while both random-slopes variants
show growing bias away from linearity (Results, "Bias"; Table 1).

This conclusion aligns with the regulatory perspective. The
FDA guidance notes that slope analyses may be considered
when 'the course of the disease is predictable and
approximately linear' [@fda2018guidance]. The EMA guideline
similarly conditions slope-based claims on verifiable
linearity [@ema2018guideline]. The ICH E9(R1) estimand
framework reinforces that the estimand (slope vs.
final-visit) should be chosen on scientific grounds before
the statistical model is specified [@ich2019e9r1].

The results from recent Phase 3 trials are instructive.
Visual inspection of CDR-SB trajectories in CLARITY-AD
[@vandyck2023lecanemab] and TRAILBLAZER-ALZ 2
[@sims2023donanemab] suggests that the treatment effect
separation is not perfectly linear. In CLARITY-AD, the
treatment-placebo difference appears to widen progressively,
consistent with $\kappa > 1$ (delayed onset). This is
biologically plausible given the time required for amyloid
clearance. The proportional MMRM approach of
@wang2022proportional offers an intermediate option that
assumes the treatment effect is proportional to the placebo
trajectory without requiring linearity per se.

## Limitations

This study has several limitations beyond the estimator issue
discussed above.

First, we now include a null scenario ($\delta = 0$; Results,
"Type I error under the null") confirming that all three models
control type I error near the nominal 0.05 level, so the power
differences reported here are not an artifact of differential
false-positive rates. We did not, however, simulate a transient
early treatment effect that vanishes by month 18 ($\delta = 0$
at $t_{\max}$ with $g(t) \not\equiv 0$ at interim visits), which
is itself a form of nonlinearity that the random-slopes model
could handle particularly poorly; this remains deferred to
future work.

Second, the design uses a single sample size (200 per arm), a
single effect size, and a single covariance structure, with
$n_{\mathrm{sim}} = 1000$ replicates (below the
$n_{\mathrm{rep}} \geq 2{,}066$ derived in Methods for the
targeted power MCSE). At 200 per arm the powers observed here
(0.54-0.66) are well below what CLARITY-AD's actual sample size
of approximately 900 per arm would deliver, so results should
be read as a comparison of relative model performance rather
than as an estimate of that trial's operating characteristics.

Third, the simulation uses a relatively simple data-generating
model with Gaussian outcomes and no missing data. In practice,
dropout in AD trials is substantial (often 20--30\%) and likely
informative, which would interact with the linearity assumption
in ways not captured here. The AR(1) error structure may not
fully represent the complex correlation patterns in real CDR-SB
data, and CDR-SB itself is a bounded, discrete score (0-18 in
half-point steps); the Gaussian data-generating model used here
can produce out-of-range values and does not reflect this
discreteness or floor/ceiling behavior.

Fourth, neither model adjusts for baseline CDR-SB (Methods,
"Analysis models"), which departs from standard
baseline-adjusted change-from-baseline MMRM practice
[@mallinckrodt2008recommendations] and was a deliberate,
disclosed simplification rather than an oversight; a
baseline-adjusted variant of both models is future work.

Fifth, 95\% CI coverage is computed using the normal critical
value 1.96 for both models (`summarize_results()`), while
$p$-values for the random-slopes model come from a $t$
reference distribution and for the categorical MMRM from a
Satterthwaite reference distribution; for internal consistency
each model's coverage should use its own critical value, which
would very slightly widen the random-slopes model's reported
intervals.

Future work might incorporate realistic dropout mechanisms, a
second sample size near 900 per arm with $n_{\mathrm{rep}} \geq
2{,}100$ (matching CLARITY-AD; still deferred, see the
Reproducibility statement), a transient-effect scenario, and
sensitivity analyses under the ICH E9(R1) framework.

# Future research

1. **Incorporate dropout.** Extend the simulation to include
   informative dropout mechanisms, evaluating how the
   interaction between nonlinearity and missingness affects
   bias and coverage under MAR and MNAR assumptions.

2. **Additional covariance structures.** Compare unstructured,
   Toeplitz, and heterogeneous AR(1) covariance models
   within the categorical-time MMRM to determine whether
   covariance misspecification compounds the effects of mean
   structure misspecification.

3. **Proportional models; design expansion.** Evaluate the
   proportional MMRM [@wang2022proportional] as an intermediate
   model that borrows strength across time without full
   linearity. The constrained longitudinal data analysis (cLDA)
   variant of the random-slopes model (Methods; Results,
   "Estimator definition check") has now been evaluated at full
   scale and integrated into Table 1; remaining design-expansion
   work is a second sample size near 900 per arm with MAR
   dropout and $n_{\mathrm{rep}} \geq 2{,}100$ (see the
   Reproducibility statement).

4. **Nonlinear mixed-effects models.** Investigate whether
   fitting a power-function or sigmoidal treatment effect
   model can recover the true $\kappa$ and provide unbiased
   estimates even under strong nonlinearity, following the
   approach of @wang2024nonlinear.

5. **Bayesian MMRM.** Evaluate whether Bayesian approaches
   with weakly informative priors on the treatment effect
   shape can adaptively borrow information across time points
   without the rigidity of the linear assumption.

6. **Software development.** Develop an R package that
   implements the simulation framework and provides
   diagnostic tools for assessing linearity of the treatment
   effect in observed trial data, building on the `mmrm`
   package infrastructure [@sabanesbove2024mmrm].

7. **Applied re-analysis.** Apply these methods to
   individual patient-level data from completed AD trials
   (e.g., ADNI placebo arms) to estimate the empirical
   $\kappa$ for different endpoints and disease stages
   [@jamalian2023modeling].

# Reproducibility statement {-}

This simulation follows the ADEMP reporting framework of
Morris, White, and Crowther [-@morris2019using]. RNGkind is
pinned to `"L'Ecuyer-CMRG"`. `analysis/scripts/run_sim_08.R`
(main design, seed `20260310`) and
`analysis/scripts/run_sim_08_null.R` (null scenario, seed
`20260320`, chosen distinct from the main design's seed so the
two RNG streams never overlap) each set their seed once before
simulation begins, not inside the per-replicate worker; both
call the parallel driver `run_simulation_parallel()` in
`analysis/scripts/sim_mmrm_linearity.R`, which assigns each
parallel worker its own reproducible `L'Ecuyer-CMRG` substream
via `parallel::nextRNGStream()` (documented in that function's
source comments); `summarize_results()` reports the convergence
rate and Monte Carlo SE for every performance measure (Table 2,
Table 3). Because work is partitioned across cores rather than
run on a single stream, results are reproducible at a fixed
seed and `mc.cores` but are not bit-identical to a serial
`run_simulation()` run at the same seed. An earlier internal
audit of this pipeline against the same framework, which
predates these fixes, is archived at
`docs/morris-audit-2026-04-17.md` and is not part of this
manuscript.

The results in this report are loaded from
`analysis/data/derived_data/sim_08.rds`
($n_{\mathrm{rep}} = 1000$ per $\kappa$, seven values of
$\kappa$, all three estimators) and
`analysis/data/derived_data/sim_08_null.rds`
($n_{\mathrm{rep}} = 1000$, $\delta = 0$, all three estimators).
To regenerate them:

```
Rscript analysis/scripts/run_sim_08.R
Rscript analysis/scripts/run_sim_08_null.R
```

One item remains open, deferred beyond this revision's time
budget:

1. **Design expansion.** A second sample size near 900 per arm
   (matching CLARITY-AD), MAR dropout, and $n_{\mathrm{rep}}
   \geq 2{,}100$ (Limitations; Future research) would require
   extending `simulate_one_trial()` with a dropout mechanism (not
   currently implemented) and calling `run_simulation_parallel()`
   with `n_per_arm = 900`; on this hardware (8 cores), extrapolating
   from the main design's wall-clock time, this is estimated to
   take on the order of an hour or more per design point and was
   out of scope for this revision. No numbers for this extension
   are reported in this manuscript.

# References
