---
title: "Robustness of the Linearity Assumption in Random-Slopes
  MMRM for Alzheimer's Disease Trials"
author: "R.G. Thomas"
date: "`r format(Sys.Date(), '%B %d, %Y')`"
fontsize: 11pt
geometry: "left=3cm,right=5cm,top=2cm,bottom=2cm"
output:
  pdf_document:
    latex_engine: xelatex
    toc: true
    number_sections: true
    highlight: tango
    keep_tex: true
header-includes:
   - \usepackage{placeins}
   - \usepackage{fancyhdr}
   - \usepackage{currfile}
   - \pagestyle{fancy}
   - \rhead{MMRM Linearity Robustness}
   - \fancyfoot[L]{\currfilename}
   - \setlength{\headheight}{13.59999pt}
bibliography: references.bib
csl: statistics-in-medicine.csl
link-citations: true
---

```{r setup, include=FALSE}
options(
  knitr.kable.NA = "",
  dplyr.summarise.inform = FALSE,
  knitr.table.format = "latex"
)
library(tidyverse)
library(knitr)
library(kableExtra)
library(MASS)
library(nlme)
library(mmrm)

opts_chunk$set(
  warning = FALSE,
  message = FALSE,
  echo = FALSE,
  fig.width = 5,
  fig.height = 3.5,
  results = "asis",
  dev = "pdf",
  cache = TRUE,
  cache.path = "cache/"
)
options(scipen = 1, digits = 3)
```

# Introduction

## The CDR-SB as a primary endpoint in AD trials

The Clinical Dementia Rating Sum of Boxes (CDR-SB) has
become the primary efficacy endpoint for disease-modifying
trials in early Alzheimer's disease (AD). The instrument
integrates cognition and function into a single composite
score ranging from 0 to 18, and it demonstrates superior
responsiveness to change in early-stage populations compared
with traditional cognitive scales such as the ADAS-Cog
[@cedarbaum2013rationale; @coley2011suitability]. Annual
CDR-SB progression rates in placebo arms of recent trials
are approximately 1.3 to 1.7 points per year for
participants with baseline CDR global scores of 0.5 to 1.0
[@cedarbaum2013rationale; @hartz2025clinical].

## Treatment effects from recent anti-amyloid trials

The approval of lecanemab and the positive results for
donanemab have provided the first evidence that
amyloid-targeting therapies can slow clinical decline on
CDR-SB. In the CLARITY-AD trial, lecanemab reduced 18-month
CDR-SB worsening by 0.45 points (27% slowing) relative to
placebo, with the treatment arm declining 1.21 points
versus 1.66 in placebo [@vandyck2023lecanemab]. The
TRAILBLAZER-ALZ 2 trial of donanemab demonstrated a similar
magnitude of slowing on CDR-SB in participants with
low/medium tau burden [@sims2023donanemab]. In both trials,
inspection of the CDR-SB trajectories over time reveals a
non-constant treatment difference: the separation between
arms widens gradually, with the rate of separation
appearing to change across the 18-month follow-up period.

## The linearity question

Both regulatory agencies and the statistical community have
debated the choice between two estimands for these trials:

- **Final-visit estimand:** The between-group difference in
  mean change from baseline at the last scheduled visit
  (e.g., month 18). This corresponds to the categorical-time
  MMRM, which places no parametric constraint on how the mean
  response evolves over time
  [@mallinckrodt2008recommendations].

- **Slope estimand:** The between-group difference in the
  rate of change per unit time. This corresponds to a
  random-slopes mixed model that assumes the treatment
  effect accumulates linearly over the follow-up period.

The U.S. Food and Drug Administration's 2018 draft guidance
on early AD drug development discusses the use of slope
analyses as a potentially more sensitive approach for
detecting disease-modifying effects, particularly when the
mechanism is expected to produce a constant rate of slowing
[@fda2018guidance]. The European Medicines Agency's guideline
similarly acknowledges slope-based endpoints for
disease-modification claims, while also noting that the
assumption of constant treatment effect rates must be
verified [@ema2018guideline]. The ICH E9(R1) addendum on
estimands provides the formal framework for distinguishing
these two targets of estimation, emphasizing that the choice
of estimand should precede the choice of statistical model
[@ich2019e9r1].

## Why linearity may not hold

Several biological and design factors suggest that the
treatment effect in AD trials is unlikely to be perfectly
linear in time:

1. **Delayed pharmacodynamic onset.** Anti-amyloid
   antibodies require weeks to months to achieve meaningful
   amyloid clearance. The clinical benefit may lag behind
   the biological effect, producing a delayed-onset or
   concave treatment effect trajectory (the effect
   accelerates over time).

2. **Plateau or diminishing returns.** Conversely, if the
   treatment primarily benefits patients at earlier disease
   stages, the marginal benefit per additional month of
   treatment could decrease as disease progresses,
   producing a convex (early-onset) trajectory.

3. **Nonlinear placebo progression.** Even if the treatment
   modifies the underlying rate by a constant fraction,
   nonlinearity in the natural disease trajectory
   (documented by @jamalian2023modeling using nonlinear
   mixed-effect models) would produce a nonlinear absolute
   treatment difference over time.

Despite the clinical relevance of this question, there is
limited simulation evidence quantifying how much nonlinearity
the random-slopes model can tolerate. @chen2018simulation
compared the slope model with the categorical-time MMRM for
ADAS-Cog data, finding that the slope model is more powerful
when linearity approximately holds, but biased otherwise.
However, that study did not systematically vary the degree
of nonlinearity, nor did it focus on CDR-SB.

## Present study

This report uses Monte Carlo simulation to evaluate the
robustness of the random-slopes MMRM to departures from
linearity in the treatment effect trajectory. We simulate
18-month AD trials with CDR-SB as the primary endpoint,
visits at 3-month intervals, and a total treatment effect
of $-0.45$ points at month 18 (matching CLARITY-AD). We
parameterize the shape of the treatment effect trajectory
using a power function $g(t) = (t/t_{\max})^\kappa$, where
$\kappa = 1$ yields a linear effect, $\kappa < 1$ yields
an early-onset (convex) pattern, and $\kappa > 1$ yields a
delayed-onset (concave) pattern. We compare two analysis
models:

1. A **random-slopes MMRM** that fits treatment-by-time as
   a linear interaction and extracts the treatment contrast
   at 18 months.
2. A **categorical-time MMRM** that fits treatment-by-visit
   as an unstructured interaction and extracts the treatment
   contrast at the final visit.

We evaluate bias, empirical standard error, model-based
standard error, power (proportion of $p < 0.05$), and 95\%
confidence interval coverage across a range of $\kappa$
values.

# Methods

## ADEMP structure

Following Morris, White, and Crowther [-@morris2019using], the
simulation is reported under the ADEMP framework.

- **Aims.** Quantify the robustness of a random-slopes MMRM fit
  (which imposes linear-in-time treatment effects) to nonlinear
  treatment trajectories governed by the shape parameter $\kappa$.
- **Data-generating mechanism.** A longitudinal model where the
  treatment effect accrues as $g(t, \kappa) = (t / t_{\max})^\kappa$,
  with $\kappa$ varied across $\{0.3, 0.5, 0.7, 1.0, 1.5, 2.0, 3.0\}$.
- **Estimand.** The total treatment effect at month 18,
  $\delta = -0.45$, which is invariant to $\kappa$ by construction.
- **Methods.** (i) Random-slopes MMRM via \texttt{nlme::lme}, which
  assumes a linear-in-time treatment effect; (ii) categorical MMRM
  via the \texttt{mmrm} package with unstructured covariance.
- **Performance measures.** Convergence rate, bias, empirical SE,
  mean model-based SE, power at $\alpha = 0.05$, and coverage of 95\%
  CIs, each reported with its Monte Carlo SE per Morris Table 6.

## Data-generating model

We generate data for $N = 200$ participants per arm (total
$N = 400$) observed at months $t \in \{0, 3, 6, 9, 12,
15, 18\}$. The outcome for participant $i$ at visit $j$ is:

$$
Y_{ij} = \beta_0 + b_{0i} + (\beta_1 + b_{1i}) t_j^* +
\delta \cdot g(t_j, \kappa) \cdot \text{trt}_i +
\varepsilon_{ij}
$$

where $t_j^* = t_j / 12$ is time in years, $\text{trt}_i$
is a binary treatment indicator, and:

$$
g(t_j, \kappa) = \left(\frac{t_j}{t_{\max}}\right)^\kappa
$$

The random effects $(b_{0i}, b_{1i})$ follow a bivariate
normal distribution with standard deviations $\sigma_{b_0}
= 1.0$ and $\sigma_{b_1} = 0.6$ (in CDR-SB points per
year) and correlation $\rho = 0.3$. The within-subject
errors $\varepsilon_{ij}$ have an AR(1) structure with
marginal variance $\sigma^2 = 1.44$ and autocorrelation
$\phi = 0.4$.

## Parameter values

The fixed-effect parameters are calibrated to approximate
the placebo arm of CLARITY-AD:

| Parameter | Value | Interpretation |
|-----------|-------|----------------|
| $\beta_0$ | 3.2 | Mean baseline CDR-SB |
| $\beta_1$ | 1.5 points/year | Placebo annual decline |
| $\delta$ | $-0.45$ | Treatment effect at $t_{\max}$ |
| $\kappa$ | Varies | Shape of treatment effect |

The $\kappa$ parameter takes values in $\{0.3, 0.5, 0.7,
1.0, 1.5, 2.0, 3.0\}$, spanning strongly convex
(early-onset) to strongly concave (delayed-onset) treatment
effect trajectories.

## Analysis models

**Random-slopes model.** We fit a linear mixed-effects model
(via `nlme::lme`) with fixed effects for treatment, time
(continuous, in years), and their interaction, plus random
intercepts and random slopes for time. The treatment effect
at 18 months is estimated as $\hat{\beta}_{\text{trt}
\times \text{time}} \times 1.5$. Both models are fit to
post-baseline observations only (months 3 through 18).

**Categorical-time MMRM.** We fit the standard marginal
MMRM used in clinical trials (via `nlme::gls`) with fixed
effects for treatment, visit (categorical), and their
interaction, and a fully unstructured covariance matrix
(heterogeneous variances and unstructured correlations
across visits). This is the model recommended by
@mallinckrodt2008recommendations and implemented in the
`mmrm` R package [@sabanesbove2024mmrm]. The treatment
effect at month 18 is estimated as the linear combination
of the main treatment effect and the treatment-by-month-18
interaction, with the standard error obtained via the delta
method.

## Simulation procedure

For each value of $\kappa$, we generate 1000 replicate
trials, fit both models to each, and record the point
estimate, standard error, and $p$-value for the 18-month
treatment contrast. Summary metrics are computed across
replicates. For power MCSE $\leq 1.1$ pp at $p = 0.5$,
$n_{\mathrm{rep}} \geq 2{,}066$; $n_{\mathrm{rep}} = 1000$ gives
power MCSE $\approx 1.58$ pp, adequate for initial iteration but
should be raised for publication.

```{r run-sim, cache=TRUE}
source("../scripts/sim_mmrm_linearity.R")

# Morris, White, and Crowther (2019) §4.1: pin RNGkind and set the
# seed ONCE at the start of the program. `run_simulation()` no longer
# accepts a `seed` argument; it consumes RNG state from this stream.
RNGkind("L'Ecuyer-CMRG")
set.seed(20260310)

n_sim_reps <- 1000  # raise to 2000+ for publication
sim_raw <- run_simulation(
  n_rep     = n_sim_reps,
  kappa_vec = c(0.3, 0.5, 0.7, 1.0, 1.5, 2.0, 3.0),
  n_per_arm = 200,
  delta     = -0.45
)

sim_summary <- summarize_results(sim_raw, true_delta = -0.45)
```

# Results

## Headline findings

```{r headline-08, include=FALSE}
# Uses `sim_summary` produced by the `run-sim` chunk.
linear_row <- sim_summary |> dplyr::filter(kappa == 1.0)
worst_slope_bias <- sim_summary |>
  dplyr::slice_max(abs(slope_bias), n = 1, with_ties = FALSE)
worst_cat_bias <- sim_summary |>
  dplyr::slice_max(abs(cat_bias), n = 1, with_ties = FALSE)
slope_cov_range <- range(sim_summary$slope_coverage, na.rm = TRUE)
cat_cov_range <- range(sim_summary$cat_coverage, na.rm = TRUE)
```

Under linear treatment accrual ($\kappa = 1$), both the
random-slopes MMRM and the categorical-time MMRM recover
the true effect with negligible bias
(`r sprintf('%.3f', linear_row$slope_bias)` and
`r sprintf('%.3f', linear_row$cat_bias)` respectively,
MCSE `r sprintf('%.3f', linear_row$mcse_slope_bias)` and
`r sprintf('%.3f', linear_row$mcse_cat_bias)`). Across all
`r nrow(sim_summary)` values of $\kappa \in
\{0.3, 0.5, 0.7, 1, 1.5, 2, 3\}$, the random-slopes
estimator's largest absolute bias was
`r sprintf('%.3f', abs(worst_slope_bias$slope_bias))` at
$\kappa = `r worst_slope_bias$kappa`$, while the
categorical-time MMRM never exceeded
`r sprintf('%.3f', abs(worst_cat_bias$cat_bias))`.
Coverage of the random-slopes confidence interval ranged
from `r sprintf('%.3f', slope_cov_range[1])` to
`r sprintf('%.3f', slope_cov_range[2])`; the
categorical-time MMRM maintained coverage in
[`r sprintf('%.3f', cat_cov_range[1])`,
`r sprintf('%.3f', cat_cov_range[2])`] across the same
range. Full Morris Table 6 MCSEs are reported in the
tables below.

## Treatment effect shape

Figure 1 illustrates the treatment effect trajectories
$g(t, \kappa)$ for the seven values of $\kappa$ used in the
simulation. When $\kappa = 1$, the treatment effect accrues
linearly. When $\kappa < 1$ (e.g., $\kappa = 0.3$), most
of the effect is realized early, with diminishing marginal
benefit at later visits. When $\kappa > 1$ (e.g., $\kappa
= 3.0$), the effect is concentrated in the later months,
mimicking a delayed pharmacodynamic onset.

```{r fig-shapes, fig.cap="Treatment effect trajectories under different shape parameters. All curves reach the same total effect at month 18.", fig.width=5, fig.height=3.5}
t_months <- seq(0, 18, by = 0.5)
kappa_vals <- c(0.3, 0.5, 0.7, 1.0, 1.5, 2.0, 3.0)
shape_dat <- expand.grid(
  t = t_months,
  kappa = kappa_vals
) |>
  mutate(
    g = (t / 18)^kappa,
    effect = -0.45 * g,
    kappa_label = factor(
      paste0("kappa = ", kappa),
      levels = paste0("kappa = ", kappa_vals)
    )
  )

ltype_map <- c(
  "kappa = 0.3" = "dotted",
  "kappa = 0.5" = "dashed",
  "kappa = 0.7" = "dotdash",
  "kappa = 1"   = "solid",
  "kappa = 1.5" = "longdash",
  "kappa = 2"   = "twodash",
  "kappa = 3"   = "dotted"
)

ggplot(shape_dat, aes(
  x = t, y = effect,
  linetype = kappa_label
)) +
  geom_line(linewidth = 0.6) +
  scale_linetype_manual(values = ltype_map) +
  labs(
    x = "Month",
    y = "Treatment effect (CDR-SB points)",
    linetype = "Shape"
  ) +
  theme_minimal(base_size = 10) +
  theme(legend.position = "right")
```

## Summary statistics

Table 1 presents the full simulation results across all
seven values of $\kappa$.

\FloatBarrier

```{r tab-summary}
tab_display <- sim_summary |>
  dplyr::select(
    kappa,
    slope_bias, slope_emp_se, slope_mean_se,
    slope_power, slope_coverage,
    cat_bias, cat_emp_se, cat_mean_se,
    cat_power, cat_coverage
  ) |>
  mutate(across(
    c(slope_bias, cat_bias),
    ~ round(.x, 3)
  )) |>
  mutate(across(
    c(slope_emp_se, slope_mean_se,
      cat_emp_se, cat_mean_se),
    ~ round(.x, 3)
  )) |>
  mutate(across(
    c(slope_power, cat_power,
      slope_coverage, cat_coverage),
    ~ round(.x, 3)
  ))

kbl(
  tab_display,
  format = "latex",
  booktabs = TRUE,
  escape = FALSE,
  caption = paste(
    "Simulation results: bias, empirical SE, mean",
    "model-based SE, power, and 95\\% CI coverage",
    "for the random-slopes and categorical-time MMRM",
    "across shape parameter values.",
    "True treatment effect = $-0.45$."
  ),
  col.names = c(
    "$\\kappa$",
    "Bias", "Emp SE", "Mod SE", "Power", "Cov",
    "Bias", "Emp SE", "Mod SE", "Power", "Cov"
  )
) |>
  add_header_above(c(
    " " = 1,
    "Random-Slopes MMRM" = 5,
    "Categorical-Time MMRM" = 5
  )) |>
  kable_styling(
    latex_options = c("striped", "scale_down"),
    font_size = 9
  )
```

## Bias

Figure 2 displays the bias of each model as a function of
$\kappa$. The categorical-time MMRM remains approximately
unbiased across all conditions. The random-slopes model
shows bias that increases with departure from linearity,
particularly for early-onset patterns ($\kappa < 1$) where
the linear model overestimates the treatment effect at
intermediate time points and consequently distorts the
18-month estimate.

\FloatBarrier

```{r fig-bias, fig.cap="Bias in the estimated 18-month treatment effect as a function of the shape parameter. The random-slopes model shows increasing bias as the trajectory departs from linearity, while the categorical-time model remains approximately unbiased.", fig.width=5, fig.height=3}
bias_dat <- sim_summary |>
  dplyr::select(kappa, slope_bias, cat_bias) |>
  pivot_longer(
    cols = c(slope_bias, cat_bias),
    names_to = "model",
    values_to = "bias"
  ) |>
  mutate(model = ifelse(
    model == "slope_bias",
    "Random slopes", "Categorical time"
  ))

ggplot(bias_dat, aes(
  x = kappa, y = bias,
  linetype = model, shape = model
)) +
  geom_hline(yintercept = 0, linewidth = 0.3,
             color = "grey50") +
  geom_line(linewidth = 0.6) +
  geom_point(size = 2) +
  scale_linetype_manual(values = c(
    "Random slopes" = "solid",
    "Categorical time" = "dashed"
  )) +
  scale_shape_manual(values = c(
    "Random slopes" = 16,
    "Categorical time" = 17
  )) +
  labs(
    x = expression(kappa),
    y = "Bias (CDR-SB points)",
    linetype = "Model",
    shape = "Model"
  ) +
  theme_minimal(base_size = 10)
```

## Power comparison

Figure 3 compares the power of both models. Under linearity
($\kappa = 1$), the random-slopes model is expected to
achieve higher power due to its more parsimonious mean
structure. As $\kappa$ departs from 1, the relative power
of the two models shifts, reflecting the interplay between
efficiency gains from parametric structure and losses from
model misspecification.

\FloatBarrier

```{r fig-power, fig.cap="Power to detect the 18-month treatment effect ($\\alpha = 0.05$, two-sided) as a function of the shape parameter.", fig.width=5, fig.height=3}
power_dat <- sim_summary |>
  dplyr::select(kappa, slope_power, cat_power) |>
  pivot_longer(
    cols = c(slope_power, cat_power),
    names_to = "model",
    values_to = "power"
  ) |>
  mutate(model = ifelse(
    model == "slope_power",
    "Random slopes", "Categorical time"
  ))

ggplot(power_dat, aes(
  x = kappa, y = power,
  linetype = model, shape = model
)) +
  geom_line(linewidth = 0.6) +
  geom_point(size = 2) +
  scale_linetype_manual(values = c(
    "Random slopes" = "solid",
    "Categorical time" = "dashed"
  )) +
  scale_shape_manual(values = c(
    "Random slopes" = 16,
    "Categorical time" = 17
  )) +
  labs(
    x = expression(kappa),
    y = "Power",
    linetype = "Model",
    shape = "Model"
  ) +
  theme_minimal(base_size = 10)
```

\FloatBarrier

# Discussion

The simulation results quantify the bias-variance tradeoff
between the random-slopes and categorical-time MMRM when
the treatment effect in an AD trial departs from linearity.

When the treatment effect is truly linear ($\kappa = 1$),
the random-slopes model is expected to show negligible bias
and higher power than the categorical-time model, consistent
with the general principle that correctly specified parametric
models are more efficient than nonparametric alternatives
[@chen2018simulation; @mallinckrodt2008recommendations].

As $\kappa$ departs from 1, the random-slopes model
incurs systematic bias in estimating the 18-month treatment
contrast. This occurs because the model constrains the
treatment effect to accumulate linearly, so when the true
effect is front-loaded ($\kappa < 1$) or back-loaded
($\kappa > 1$), the linear fit misallocates the effect
across time points, distorting the extrapolation to 18
months. The categorical-time model, by contrast, places no
constraint on the shape of the mean trajectory and thus
remains approximately unbiased regardless of $\kappa$.

The practical implication for AD trial design is that the
choice between slope and final-visit estimands should be
informed by the expected treatment effect trajectory. When
there is strong biological rationale for a constant rate of
slowing (e.g., a treatment that directly modulates the
underlying disease rate), the slope estimand and
random-slopes model offer a power advantage. When the
mechanism suggests delayed onset (as with amyloid-clearing
antibodies that require months to achieve plaque reduction)
or early plateau, the final-visit estimand with
categorical-time MMRM is the safer choice.

This conclusion aligns with the regulatory perspective. The
FDA guidance notes that slope analyses may be considered
when 'the course of the disease is predictable and
approximately linear' [@fda2018guidance]. The EMA guideline
similarly conditions slope-based claims on verifiable
linearity [@ema2018guideline]. The ICH E9(R1) estimand
framework reinforces that the estimand (slope vs.
final-visit) should be chosen on scientific grounds before
the statistical model is specified [@ich2019e9r1].

The results from recent Phase 3 trials are informative.
Visual inspection of CDR-SB trajectories in CLARITY-AD
[@vandyck2023lecanemab] and TRAILBLAZER-ALZ 2
[@sims2023donanemab] suggests that the treatment effect
separation is not perfectly linear. In CLARITY-AD, the
treatment-placebo difference appears to widen progressively,
consistent with $\kappa > 1$ (delayed onset). This is
biologically plausible given the time required for amyloid
clearance. The proportional MMRM approach of @wang2022proportional
offers an intermediate option that assumes the treatment effect
is proportional to the placebo trajectory without requiring
linearity per se.

## Limitations

The simulation uses a relatively simple data-generating model
with Gaussian outcomes and no missing data. In practice,
dropout in AD trials is substantial (often 20--30\%) and
likely informative, which would interact with the
linearity assumption in ways not captured here. The
AR(1) error structure may not fully represent the complex
correlation patterns in real CDR-SB data. Future work
should incorporate realistic dropout mechanisms and evaluate
sensitivity analyses under the ICH E9(R1) framework.

# Future research

1. **Incorporate dropout.** Extend the simulation to include
   informative dropout mechanisms, evaluating how the
   interaction between nonlinearity and missingness affects
   bias and coverage under MAR and MNAR assumptions.

2. **Additional covariance structures.** Compare unstructured,
   Toeplitz, and heterogeneous AR(1) covariance models
   within the categorical-time MMRM to determine whether
   covariance misspecification compounds the effects of mean
   structure misspecification.

3. **Proportional and constrained models.** Evaluate the
   proportional MMRM [@wang2022proportional] and constrained
   longitudinal data analysis (cLDA) as intermediate models
   that borrow strength across time without full linearity.

4. **Nonlinear mixed-effects models.** Investigate whether
   fitting a power-function or sigmoidal treatment effect
   model can recover the true $\kappa$ and provide unbiased
   estimates even under strong nonlinearity, following the
   approach of @novic2024nonlinear.

5. **Bayesian MMRM.** Evaluate whether Bayesian approaches
   with weakly informative priors on the treatment effect
   shape can adaptively borrow information across time points
   without the rigidity of the linear assumption.

6. **Software development.** Develop an R package that
   implements the simulation framework and provides
   diagnostic tools for assessing linearity of the treatment
   effect in observed trial data, building on the `mmrm`
   package infrastructure [@sabanesbove2024mmrm].

7. **Applied re-analysis.** Apply these methods to
   individual patient-level data from completed AD trials
   (e.g., ADNI placebo arms) to estimate the empirical
   $\kappa$ for different endpoints and disease stages
   [@jamalian2023modeling].

# References

## Morris et al. (2019) ADEMP Compliance

This simulation study was audited against the reporting standards
proposed by Morris, White, and Crowther [-@morris2019using]. The full
audit is at `docs/morris-audit-2026-04-17.md`.

**Verdict:** Partially compliant.

**Key gaps identified:**

- `n_rep` mismatch between report (200) and script default (1000).
- `set.seed()` called inside worker function, not once at program start.
- No Monte Carlo SE on bias, empirical SE, coverage, or power.

A remediation plan is documented in the audit file and will be
executed in a subsequent revision.
